---
title: "Laboratory 1"
author: "Written by Mark McAvoy"
date: "Summer 2020"
knit: (function(inputFile, encoding) { 
  out_dir <- 'Knit files (Output)';
  rmarkdown::render(inputFile,
                    encoding=encoding, 
                    output_file=file.path(dirname(inputFile), out_dir, 'Laboratory 1.html')) })
output:
  revealjs::revealjs_presentation:
    theme: beige
    highlight: pygments
    center: true
    css: styles.css

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Clear enviroment before starting 
remove(list = ls())


# The following sets your working directory to wherever this .Rmd file is saved
# note, must be included in the setup portion of the .Rmd
knitr::opts_knit$set(root.dir = getwd())
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Note: this only runs when knit, so if try running by hand it will not do anything, best to also change working directory manually
# alternatively, you may change working directory manually


# if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org") # run this the first time (only need to once), note moved tidyverse to after AER package, as it was creating some issues running AER after tidyverse
#   library(tidyverse) # includes ggplot2 and many more useful libraries
# if(!require(knitr)) install.packages("knitr", repos = "http://cran.us.r-project.org")
#   library(knitr) # knitr is used to render this R code to a LaTeX file (includes kable)
# if(!require(stargazer)) install.packages("stargazer", repos = "http://cran.us.r-project.org")
#   library(stargazer) # knitr is used to render this R code to a LaTeX file (includes kable)
# if(!require(printr)) install.packages("printr", repos = "http://cran.us.r-project.org")
# library(printr)
```

### <span style="color:deepskyblue; font-family: Times-New-Roman, Times, serif"> --- Reproducible Research with R  --- </span>
#### <span style="color:deepskyblue"> --- Chapter 3.1 - Using R: The Basics--- </span>

Please send me the code to solutions to Laboratory assignments 1-6 in either a .R file or any text document such as .pdf, .txt, .doc, etc.

## <span style="color:deepskyblue"> --- Section 3.1.1 - Objects --- </span>

Let's look at Numerics
```{r}
# Assign the value 10 into an object called Number using the assignment operator <- 
Number <- 10

# We can similarly do this with =, however it is standard to use <- when assigning an object
# in other cases we will use =, thus using <- in assigning will make reading code easier later
Number = 10

# use the print() function to display an object
print(Number)
```

## Let's look at Characters

```{r}
# Character
Words <- "Hello World"

# cat is an altenative to print
cat(Words)

# use class() to check the class of an object
class(Words)
```

## Let's look at Vectors 

```{r}
# use c() to combine elements into a vector structure
NumericVect <- c(2.8, 2, 14.8)

# Display the vector by typing just its name
NumericVect

# We can similarly make a character vector
CharacterVect <- c("Albania", "Botswana", "Cambodia")

# Display this vector by running the objects name
CharacterVect
```

## Let's look at data frames

```{r, eval = FALSE}
# combine the two vectors together into a dataframe using the function data.frame
StringNumObject <- data.frame(CharacterVect, NumericVect)

# display this dataframe
StringNumObject
```

```{r, echo = FALSE}
StringNumObject <- data.frame(CharacterVect, NumericVect)
kable(StringNumObject)
```

---

```{r, eval = FALSE}
# we can create dataframes in one step as well
# note the enter line after the , this is done to make readability easier.
OurDf <- data.frame(first_column = c('a', 'b', 'c'),
                    second_column = c(1, 2, 3))
OurDf
```

```{r, echo = FALSE}
OurDf <- data.frame(first_column = c('a', 'b', 'c'),
                    second_column = c(1, 2, 3))
kable(OurDf)
```

## <span style="color:orchid"> +++ Laboratory assignment 1-3 +++ </span>

Complete the missing parts in the following 3 questions

```{r, eval = FALSE}
# 1)
# Make a character object called HelloWrold with the string 'Hello World this is ---' where --- is your name
HelloWorld <- " "
  
# Display HelloWorld
HelloWorld

# 2)
# make a vector of the following 10 numbers, 0, 1, 1, 2, 3, 5, 8, 13, 21, 34
OurVect <- c( )

# Display this vector
OurVect

# 3)
# make a dataframe with two columns:
# the first column are 3 country names of your choosing, have this column called, 'country'
# for the second column, do a quick google search to find these countries GDP per capita in 2017 (units in trillions of USD), call this column GDP
OurFirstDf <- data.frame( )

# Display this dataframe
OurFirstDf
```

## <span style="color:deepskyblue"> --- Section 3.1.2 - Component Selection --- </span>

```{r}
# Select the NumericVect object saved in the StringNumObject using the component selector
NewNumeric <- StringNumObject$NumericVect

# R has quite a few datasets that come pre-installed in it. One of these is the Cars dataset
# we can import it into our session using the following command
attach(cars)
# not we do not see it, but it is there. We can save it into our environment by assigning it
# notice it has 2 variables, speed and distance (dist)
cars <- cars
```

## <span style="color:deepskyblue"> --- Section 3.1.3 - Subscripts --- </span>

Square brackets written after a dataframe (with no space) calls the elements of the dataframe

For example $df[1,1]$ calls the first element of the dataframe called $df$

\vspace{20pt}

Note: This will not run if you have a space between the dataframe name and the square brackets. $df \; [1,1]$ is incorrect

---

```{r, eval = FALSE}
# use head to display the first 6 rows of a dataframe
head(cars)
```

```{r, echo = FALSE}
kable(head(cars))
```

---

```{r, eval = FALSE}
# display only selected rows
cars[3:7, ]
```

```{r, echo = FALSE}
kable(cars[3:7, ])
```

---

```{r, eval = FALSE}
# or specify multiple columns in specified rows
cars[4, c("speed", "dist")]
```

```{r, echo = FALSE}
# or specify multiple columns in specified rows
kable(cars[4, c("speed", "dist")])
```

## <span style="color:deepskyblue"> --- Section 3.1.4 - Functions and Commands --- </span>

Use parentheses to call a function. For example $mean(.)$. Where we write a vector in place of the dot such as $mean(vec)$ or $mean(df\$x)$

```{r}
# use the mean function to find the mean of NumericVect
mean(x = NumericVect)

# we actually don't need to include the x = ..., just the object name inside the parethesis
mean(NumericVect)
```

## <span style="color:deepskyblue"> --- Section 3.1.5 - Arguments --- </span>

An argument is a variable called by a function.  Say for example we have the following function,

$$ f(x) = x^2  $$

Then $f(.)$ is the function and $x$ is the argument.  We can have multiple arguments, see for example, 

$$ f(x,y) = x^2 + y $$

Now the function $f(.)$ has two arguments, $x$ and $y$

## <span style="color:deepskyblue"> --- Section 3.1.6 - The workspace \& history --- </span>

We will have a full week covering the workspace and history in the third week (Chapter 4.1 in the textbook). Briefly covered here is the $ls()$ function. We will cover the save, load, read.csv, functions later

```{r}
# display all objects in your Global Environment
ls()

# note the broom icon in the environment window will remove all objects
# note your global environment is cleared upon closing or restarting your R session
```

## <span style="color:deepskyblue"> --- Section 3.1.7 - Global R options --- </span>

You can set the digits to display line-by-line using the $round()$

```{r}
options(digits = 1)
```

## <span style="color:deepskyblue"> --- Section 3.1.8 - Installing new packages and loading functions --- </span>

We'll learn installing and loading packages later

## <span style="color:orchid"> +++ Laboratory assignment 4-6 +++ </span>

Answer and display the following 3 questions

```{r}
# 4) Display the second and third rows of OurFirstDf

# 5) Display the mean of GDP per capita in OurFirstDf

# 6) Display the variance of GDP per capita in OurFirstDf

```

That's it for the first Laboratory! Enjoy the rest of your week~